# The Proposal
<!--
This is where the proposal should be outlined. 
-->

### Overview
<!--
At a high-level address what your proposal is and how it will address the problem identified. Highlight any benefits to the R Community that follow from solving the problem. This should be your most compelling section.
-->

Biostrings is a core Bioconductor package providing efficient containers for storing, manipulating, and analyzing biological sequences. Biostrings is **the** method to access biological sequence data in R--nearly every analysis working with genomic data depends on the Biostrings package to handle sequencing data. Presently, Biostrings maintenance is hindered by (i) a list of open issues on GitHub, (ii) a to do list of items developed by package contributors, and (iii) input/output functionality that is becoming outdated with newer technology. 

This project proposes to clear out this accumulated technical debt by addressing open issues, implementing robust tests for long-term sustainability, and adding features that will keep Biostrings relevant for modern sequencing technologies. For end-users, this will result in numerous bugfixes, a host of new features to support genomic analyses, and a variety of performance improvements to bolster R as one of the top programming languages for bioinformatics. For developers, this will make the Biostrings package more sustainable, allowing for more community contribution and faster bug resolution in the future.

In summary, this proposal details a new era for Biostrings. The project will transition maintenance to a new developer, and in the process, ensure the package is robust and maintainable for years to come.

### Detail
<!--
Go into more detail about the specifics of the project and it delivers against the problem.

Depending on project type the detail section should include:

 - [ ] Minimum Viable Product
 - [ ] Architecture
 - [ ] Assumptions
-->

There are three categories of changes needed to sustain the utility of Biostrings in the long-term. First, outstanding user-contributed GitHub issues must be addressed. Most of these are straightforward but require effort to complete. I anticipate up to 4 months for this first Aim. Second, current and past Biostrings developers have kept a list of to do items for the package. Prior to this proposal submission, I worked with Hervé Pagès to narrow this list to the items that are the most important items for Biostrings' long-term success. I anticipate up to 6 months will be required to complete all items on this list. Third, modern DNA sequencing technologies have advanced markedly since Biostrings was first introduced. Changes to import and export of sequences are needed to sustain the project long-term. I expect these changes will take 2 months. These three categories of changes are described in the three Aims below.

#### Aim 1: Resolving bugs and increasing robustness

<<<<<<< HEAD
The goal of this aim is the prepare the Biostrings codebase for future improvements and facilitate ongoing maintenance. Biostrings has had minimal maintenance for the past 5 years, and as a result lacks many processes that would ease continued development. The goals of the Aim are the following:
=======
The goal of this aim is to prepare the Biostrings codebase for future improvements and facilitate ongoing maintenance. Biostrings has had minimal maintenance for the past 5 years, and as a result lacks many processes that ease continued development. The goals of the Aim are the following:
>>>>>>> e9c1a52 (Update to new Aim 2)

1. Clean up outdated documentation and bug reports. Many internal documentation files and scripts have `TODO` tags or warnings that are out of date. Additionally, many bug reports on GitHub have been resolved but remain open. This makes ongoing maintenance challenging, as it requires additional developer effort to determine if a bug still exists before addressing it. This task will clean up the codebase, update outdated documentation, and clean up resolved bug reports.
2. Implement unit testing. Much of this project depends on improving and adding additional functionality to Biostrings. However, as mentioned in the Problem Definition section, Biostrings is a critical package with over a million downloads per year. As such, robust unit testing is essential to ensure that other proposed changes do not break existing functionality for end users. Robust testing suites would also make it easier for maintainers to review community-contributed code, thus making it easier to involve the community in Biostrings.
3. Resolving outstanding bug reports and old TODO items. While many bug reports on GitHub are resolved, there are many that remain unaddressed. This task will resolve any remaining bug reports. Part of this task will involve reaching out to bug reporters for additional detail, as some bug reports are years old without any updates (see [here](https://github.com/Bioconductor/Biostrings/issues/25) or [here](https://github.com/Bioconductor/Biostrings/issues/51) examples of very old bug reports). This task will also address some small high-priority tasks on the Biostrings TODO list (see below for a complete list), since they overlap with many of the GitHub issues.

The Biostrings repository contains a TODO list with many tasks, ranging from small bugs to long-term feature requests. Task 3 will address the small tasks from this list that overlap with existing GitHub issues:
 - Fixing stack overflow issues relating to `XStringSet` comparisons (mentioned in the aforementioned [Issue 51](https://github.com/Bioconductor/Biostrings/issues/51))
 - Ensure `validObject()` correctly works on `XStringSet` objects
 - Implement `patternFrequency` methods for `XStringSet` and `XStringSetViews` objects

Successful completion of this Aim will result in a cleaner Biostrings GitHub repository, resolution of outstanding user-submitted bugs, and a robust testing pipeline for future submissions. I expect this Aim to take three months; most of the GitHub issues are relatively quick to fix, so the majority of the time will be dedicated to building a robust testing infrastructure. The outstanding TODO items are also similarly quick to address--longer issues will be reserved for Aim 3.

Aim 1 is scheduled to be addressed first because it includes checks for later work and fixes for user-identified issues. I see these improvements as the highest priority for the Biostrings codebase.

#### Aim 2: Fixing String Matching

The remainder of items on the Biostrings TODO list fall into three categories: "nice to have" feature requests, low-priority issues, and problems with string matching. The first two of these categories can be delayed to future work, but the third category directly impacts users. This aim will focus on cleaning up the string matching libraries of Biostrings.

Matching strings is a common task in bioinformatics, and, while Biostrings does provide a host of tools for string matching, many of Biostrings' tools are underdeveloped and/or confusing to use. For example, the following interaction is clearly not ideal:

```{r, echo=TRUE, error=TRUE}
library(Biostrings, quietly = TRUE)
pd <- PDict(mkAllStrings(DNA_BASES, 3))
strs <- DNAStringSet(c("ATGCATGCA", "ATGATCATGA"))
matchPDict(pd, strs)
vmatchPDict(pd, strs)
```

Additionally, string matching doesn't support the widely used [IUPAC ambiguity codes](https://www.dnabaser.com/articles/IUPAC%20ambiguity%20codes.html):

```{r, echo=TRUE, error=TRUE}
library(Biostrings, quietly = TRUE)
patterns <- DNAStringSet(c("ATR", "GCY"))
pd <- PDict(patterns)
```

Searching for strings containing some ambiguities is a common task, and not supporting this functionality limits the usefulness of Biostrings for bioinformatics. This Aim will work on improving the user experience for string matching functions by adding in capabilities that users expect to find and resolving cryptic error scenarios. More specifically, this will address the following tasks:

1. Allow for variable width matching in `matchPDict`. The current implementation relies strongly on the notion of a "Trusted Band". This implementation is very difficult to understand for end-users, leading to confusion on how to use it and seemingly cryptic error messages. Trusted Bands can become a purely internal argument, removing a significant amount of user burden. This would also allow users to more easily search for variable width patterns in data, which is a common task.
2. Allow for ambiguity codes in search strings. This has a variety of use-cases--one such example is searching for a particular amino acid in a nucleotide coding sequence. Many amino acids are encoded by multiple codons, and the current implementation requires specifying all of these codons individually. For example, searching for Valine could be done by simply looking for `"GTN"` rather than `c("GTA", "GTC", "GTT", "GTG")`. This would also improve internal performance by decreasing the number of searches performed.
3. Resolve the circular error of `matchPDict` and `vmatchPDict` shown previously. This involves implementing `vmatchPDict` and removing these error messages.
4. Implement small helper functions for `PDict` objects to improve user experience. Specifically, implement `reverseComplement`, `duplicated`, and `patternFrequency` for `PDict` objects, and add the `skip.invalid.patterns` argument for `PDict` that has been promised in documentation for years. This will also add in helper functions to clarify available algorithms for `matchPDict`, which are currently relatively hard to understand.

Aim 2 is the longest aim, and is anticipated to take 6 months. While many of these tasks are nontrivial, much of the initial plans to implement them are detailed in the Biostrings TODO file. The remaining tasks in the TODO file ("nice to have" and low-priority items) will be addressed after the conclusion of this grant. See "Future Work" for more discussion on these tasks. 

Aim 2 is positioned second due to it containing large tasks that will improve user performance, but that are less critical that open bug reports. The conclusion of Aims 1 and 2 will result in a significantly improved user experience for R's bioinformatics community.

#### Aim 3: Enhancing input/output for modern sequencing technologies

Biostrings was initially developed during a time when sequencing produced megabases (~1M nucleotides) of data per run. Modern sequencing technologies easily produce gigabases (> 1B nucleotides) per run. Hence, Biostrings' input/output needs improvement to scale alongside next generation sequencing technologies. At present, Biostrings can only read and write sequences from gzip compressed files. There is an `open_input_files` function that allows reading sequences in batches, but it does not use the standard R connections interface and is cumbersomely slow on large compressed files.

To enhance Biostrings, I will add functionality for reading from standard gzip, bzip2, and xz connections in R. This will involve overhauling the `readXStringSet` functions within Biostrings. Furthermore, I will enable writing to alternative output file compression types (bzip2 and xz), while allowing for different compression levels. At present there is a `compression_level` argument in `writeXStringSet`, but it is unused by the function. I will focus on improving the speed of reading and writing from files so that large file sizes are no longer problematically slow. Collectively, these enhancements will propel Biostrings (and by extension, the R programming language itself) into the future of big biological data.

This Aim is placed last due it being oriented towards the future of Biostrings. Aim 1 addresses problems users currently have, Aim 2 addresses less critical current issues, and Aim 3 prepares Biostrings for the future.